Creating frames of reference for chert exploitation during the Late Pleistocene in Southwesternmost Iberia.

### 2022-12-22 analysis/paper/paper.Rmd
JC reviews

---
title: "Creating frames of reference for chert exploitation during the Late Pleistocene in Southwesternmost Iberia."
author:
  - name: Joana Belmiro
    email: jfbelmiro@ualg.pt
    institute: ICArEHB
    correspondence: true
  - name: João Cascalheira
    email: 
    institute: ICArEHB
    correspondence: false
  - name: Xavier Terradas
    email: 
    institute: CSIC
    correspondence: false
institute:
  - ICArEHB: Universidade do Algarve
  - CSIC: Bleh
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Southwestern Iberia has played a key role in Middle to Upper Paleolithic transitions and Human occupations. It may have allowed the establishment of eco-cultural niches, through the continuous availability of resources throughout the Late Pleistocene, such as chert raw materials. Chert procurement studies in this region have received increasing attention in the past two decades, especially those related to multi-component sites such as Vale Boi (Algarve, Portugal). However, these studies have focused on the western portion of the region and used macroscopic descriptions. For further studies regarding Human mobility and raw material procurement and use at Vale Boi, a complete reference collection is necessary. This paper focuses on the location, diversity, and availability of chert raw materials in Southern Portugal (Algarve). Through macroscopic and petrographic approaches, we characterize the geological cherts to build a frame of reference for raw material studies in the region. Our results show that there are three main chert formations within Southern Portugal. Despite the variability within formations, the cherts show sufficient differences both at a macroscopic and petrographic level to allow source attribution. These results provide the starting ground for chert raw material studies in archaeological assemblages, in eco-cultural niches such as the archaeological site of Vale Boi.

@@

Southern and Western Iberia have often been considered key areas to understanding techno-cultural transitions from the Middle Paleolithic to the end of the Upper Paleolithic. As territories located at the western tip of the European continent and with a generally stable climate even during the coldest periods punctuating the Late Pleistocene, these have been regarded as some of the most significant glacial refugia in Europe [@hewitt2000; @gómez2007; @gonzález-sampériz2010; @carvalho2022]. For this reason, Southernwestern Iberia has frequently been at the center of some of the most debated topics regarding Late Pleistocene human adaptations [@finlayson2006; @zilhão2017; @cascalheira2017]. A particularly good example is the region's role in possibly being one of the last territories to be occupied by Neanderthal populations right before their complete disappearance [@finlayson2006; @zilhão2017]. Neanderthals occupied the European continent for more than 300.000 years and are thought to have disappeared while modern humans arrived on the territory [@trinkaus1979; @mellars2004; @finlayson2006; @jennings2011; @rogers2017]. The last territory where Neanderthal populations seemed to exist was Southern Iberia, around c. 37 thousand years ago [@zilhão2017], or perhaps even earlier [@finlayson2006; @tzedakis2007; @carvalho2022] . This region is key to understanding how Neanderthals survived until such a later chronology, the degree and types of interaction those populations may have had with modern humans [@finlayson2006; @zilhão2017], and how and why they eventually went extinct [@dalen2012; @melchionna2018]. Another example is Southwestern Iberia's importance in the discussion of Upper Paleolithic technocomplexes transitions. Previous studies have discussed the territory's potential as a refugium during cold and harsh climatic conditions [@gómez2007; @rodríguez-sánchez2010; @jennings2011]. The Heinrich Event 2 (HE 2) at the onset of the Last Glacial Maximum (LGM), for example, is a period marked by important social and technological transformations. This climatic event was characterized by abrupt and drastic climatic changes that impacted human behavior all across westernmost Europe [@gamble2004; @sanchezgoñi2010]. The identification of a Proto-Solutrean phase in central and southern Portugal with a very distinct index fossil (the Vale Comprido point), and its direct association with the HE 2, put these regions amongst some of the most important case studies of how environmental dynamics have affected human adaptations during the last glacial. Other studies have expanded upon this notion of climatic refugia during harsh climatic events, to understand the Iberian Peninsula as a long-term eco-cultural refugia [@cascalheira2017]. Using this framework, this territory would consist of several ecological niches, consistently used through time, possibly due to the stability in the richness and variety of resources. This continuous use would then create long-term regional adaptative structures, which when correlated with the ecological niches, @cascalheira2017 have referred to as eco-cultural niches. In fact, it seems that a large number of caves and rockshelters in Iberia are multi-layered, giving validity to the aforementioned framework [@schmidt2012]. These eco-cultural niches provide an exceptional opportunity to understand long-term dynamics regarding biotic and abiotic resource exploitation since they can provide details on how human populations maintained or changed their adaptive systems when facing environmental changes, and cultural and social transformations or constraints. One of such possible Late Pleistocene eco-cultural niches to which this theoretical framework has been previously applied is the archaeological site of Vale Boi. This multi-component site is located in westernmost southern Iberia, in a region currently known as the Algarve, and comprises one of the most complete Upper Paleolithic chronocultural sequences of Iberia [@bicho2013; @cascalheira2013; @cascalheira2017]. Several cross-scale complex interactions have been identified, displaying resilient behaviors throughout the Upper Paleolithic maintained by their eco-cultural niche, but also adaptation behaviors motivated by niche diversity, social networks, and climatic changes [@cascalheira2017]. Some of these resilient elements are, for example, the continuous use of strategies like grease rendering and selective hunting patterns, site function, certain lithic technology patterns, and the functional specialization of lithic raw materials. The maintenancee of social networks through the identification of possible long-distance lithic raw materials has also been suggested [@pereira2016]. The identification of these patterns has been reliant on the great amount of studies that have originated from the archaeological site of Vale Boi [i.e., @bicho2012; @bicho2017; @pereira2016; @manne2012]. A large portion of these studies has focused on lithic technology [@cascalheira2010; @marreiros2015; @belmiro2020], and unlike other regions of Portugal [i.e., @aubry2004; @aubry2009; @aubry2022; @pereira2022; @pereira2021; @costa2022; @jordão2022], archaeological studies focusing on raw materials, and especially chert, have been more scarce, focusing mostly on macroscopic preliminary results or the differentiation between possibly local and non-local sorts of raw materials. Despite the scarceness of these studies, chert played a significant role in early and later prehistory at Vale Boi [@cascalheira2010; @marreiros2009; @belmiro2020], and across most of Prehistory in Southern Portugal [@zambujo1998; @bicho2003]. As an essential part of late Pleistocene hunter-gatherer adaptations, lithic raw materials have the potential to provide insights into the adaptive strategies of those populations [@féblot-augustins1993; @kuhn1995; @mellars1996], mostly regarding land-use, technological organization, but also cultural and social interactions. In fact, the selection and procurement of raw material has been suggested as a key stage for the technological organization of hunter gatherer groups [@bamforth1997]. Thus, changes in the frequencies of raw materials within the archaeological records may provide evidence for changes in that organization, or even resilience of specific choices, all of which can reflect culturally transmitted preferences within a group of hunter gatherers [@brown1999]. Several models, both formal and informal, have shown that changes of raw material in the archaeological record may reflect different procurement strategies [@binford1985; @binford1979; @gould1985; @gouldsaggers_1985], either related to mobility strategies [@ambrose1990; @kuhn2004; @kuhn1991; @mccall2007] or changes in the availability of raw materials, possibly related to environmental change [@brantingham2003; @brown1999]. Raw materials are also closely linked to the tools produced with them -- the costs of manufacturing technology can be related to the availability of raw materials in the landscape [@torrence1983; @bousman1993; @parry1987; @mackay2011], or the characteristics of the raw materials which might relate to functionality [@mackay2008; @minichillo2006; @stout2002], through the choice of raw materials for specific technologies [@brown1999]. Furthermore, changes in the raw materials can also be related to the establishment of social networks [@whallon2006] and the horizontal transmission of preferences through trade [@brown1999]. In order to explore the aforementioned theories and models to an archaeological site, especially when trying to understand land-use and the adaptive strategies of an eco-cultural niche like Vale Boi, it is necessary to know the landscape and the raw material sources available in the territory. To distinguish between local and non-local raw materials, first it is necessary to understand the characteristics of the local sources and establish a comparative database which can be used for the analysis of an assemblage. This is true for most of the informal models which try to understand why raw materials change in the archaeological record. Furthermore, formal models focusing on raw material use posit that it is necessary to have detailed sourcing data in order to adequately apply the models to a region or an archaeological assemblage [@surovell2009; @pop2015]. In fact, the creation of such a database is a starting point for most raw material-focused studies and has been applied all across the world, independent of the studied chronology [i.e., @luedtke1992; @bustillo2009; @driscoll2016; @terradas2017; @ekshtain2022]. In southern Portugal previous efforts have been made to create such a database. The work of @verissimo2004, albeit focusing on the occurrence of chert solely in western Algarve, provided the initial basis for comparative studies with the assemblages from Vale Boi based on macroscopic approaches. The creation of LusoLit, a lithotheque currently hosted at the University of Algarve [@pereirab_presenting_2016], and the collection of samples from several outcrops in the region provided a new leap in the study of chert in the region. A few geological studies have also contributed to understand the availability and characteristics of chert in southern Portugal [i.e., @rocha1976; @ribeiro2005]. Nevertheless, these studies are often unpublished or comprehend answers to geological questions, which hamper the comparative use of the data with archaeological assemblages. Furthermore, a large portion of chert-bearing outcrops in southern Portugal remains unstudied, both by archaeologists and geologists. As such, further analysis of the overall variability, location and availability of chert nodules in the Algarve is necessary to start testing behavioural models of land-use and abiotic resource exploitation. Given the potential for a in-depth raw material study at a possible Late Pleistocene eco-cultural niche such as Vale Boi, in this study we explore the location, diversity and availability of chert raw materials in the southernmost region of Portugal -- the Algarve. Our main goal is to establish a reference for chert raw materials in an understudied region in what regards to lithic materials, that can be used for future studies addressing chert exploitation in the Algarve and elsewhere. This includes the development of a methodological approach adapted to the study area, by testing the potential of macroscopic and petrographic approaches for the characterisation of regional cherts. The data presented here is integrated in an online lithotheque (LusoLit) that is now freely available and can be built upon and improved as new data becomes available.

@@

In this study we identified several sources of chert nodules in southern Portugal and characterized the regional cherts, which were of critical importance for hunter-gatherer communities during the Late Pleistocene. For this we applied a multilayered raw material analysis approach, composed of macroscopic and petrographic analyses. The results show the presence of four different chert formations, dispersed in western and eastern Algarve. Within formations, there is variability in the nodules and the outcrops. There are however identifiable macroscopic and petrographic differences between formations which allow their distinction. Although the petrographic analysis is essential to identify the fossils present in the chert, a macroscopic approach seems to be pertinent for a quick and inexpensive analysis to distinguish between cherts of the different formations. The presence of chert sources in the Algarve region, with distinguisheable characteristics between formations, which may be analysed preliminarily through macroscopic approaches, shows the potential for chert raw material studies of archaeological sites in this key area. Further steps in our study will include the use of the data gathered in the present study and the completed LusoLit lithotheque to study chert use from multilayered sites with Upper Paleolithic chronologies such as Vale Boi, and participate in the discussion of human adaptations throughout the Late Pleistocene. Furthermore, future approaches include the use of geochemical methods to further characterize these cherts and the integration of the resulting data in the online database.

@@

We would like the thank the entities that funded the current research: Fundação para a Ciência e a Technologia (FCT) and ICArEHB (Interdisciplinary Center for Archaeology and the Evolution of Human Behavior). Furthermore, we want to thank Dr. Telmo Pereira and Dr. Carlos Ribeiro for allowing us to consult their previous research and for their support. Finally, we'd like to thank Jack Acres for the IT support essential for the creation of the online LusoLit database.

>>>>

---
title: "Creating frames of reference for chert exploitation during the Late Pleistocene in Southwesternmost Iberia."
author:
  - name: Joana Belmiro
    email: jfbelmiro@ualg.pt
    institute: ICArEHB
    correspondence: true
  - name: Xavier Terradas
    email: 
    institute: CSIC
    correspondence: false
  - name: João Cascalheira
    email: 
    institute: ICArEHB
    correspondence: false
institute:
  - ICArEHB: Interdisciplinary Center for Archaeology and the Evolution of Human Behavior, University of Algarve
  - CSIC: Consejo Superior de Investigaciones Científicas
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Southwestern Iberia has played a key role in characterizing Late Pleistocene human ecodynamics. Among other aspects of human behavior, chert procurement and use studies in this region have received increasing attention in the past two decades, especially focusing on the sites showing repeated human occupation, such as the case of Vale Boi (Southern Portugal). However, these studies have been very limited in their geographical scope, and mostly focused on brief macroscopic descriptions of the raw materials. To further our knowledge of the relationship between regional availability of raw materials and its impact on human adaptations and mobility, a more detailed approach to characterizing geological sources is needed. This paper characterizes chert raw materials location, diversity, and availability in a geologically well-defined region of southern Portugal - the Algarve. Through macroscopic and petrographic approaches, we provide a detailed characterization of geological chert sources to build a frame of reference for chert exploitation in the region. Our results show that there are three main chert formations in Algarve, and that despite the within-source variability, sufficient differences at macroscopic and petrographic levels are present to allow clear source attribution. These results provide a baseline for raw material studies in archaeological assemblages across southwestern Iberia, that will be essencial to further characterize the dynamics of human behavior in some of the most important eco-cultural niches.

@@

Southern and Western Iberia have often been considered key areas to understanding techno-cultural transitions from the Middle Paleolithic to the end of the Upper Paleolithic. As territories located at the western tip of the European continent and with a generally stable climate even during the coldest periods punctuating the Late Pleistocene, these have been regarded as some of the most significant glacial refugia in Europe [@hewitt2000; @gómez2007; @gonzález-sampériz2010; @carvalho2022]. For this reason, Southernwestern Iberia has frequently been at the center of some of the most debated topics regarding Late Pleistocene human adaptations [@finlayson2006; @zilhão2017; @cascalheira2017]. A particularly good example is the region's role in possibly being one of the last territories to be occupied by Neanderthal populations right before their complete disappearance [@finlayson2006; @zilhão2017]. Neanderthals occupied the European continent for more than 300.000 years and are thought to have disappeared while modern humans arrived on the territory [@trinkaus1979; @mellars2004; @finlayson2006; @jennings2011; @rogers2017]. The last territory where Neanderthal populations seemed to exist was Southern Iberia, around c. 37 thousand years ago [@zilhão2017], or perhaps even earlier [@finlayson2006; @tzedakis2007; @carvalho2022]. This region is key to understanding how Neanderthals survived until such a later chronology, the degree and types of interaction those populations may have had with modern humans [@finlayson2006; @zilhão2017], and how and why they eventually went extinct [@dalen2012; @melchionna2018]. Another example is Southwestern Iberia's importance in the discussion of Upper Paleolithic technocomplexes transitions. Previous studies have discussed the territory's potential as a refugium during cold and harsh climatic conditions [@gómez2007; @rodríguez-sánchez2010; @jennings2011]. The Heinrich Event 2 (HE 2) at the onset of the Last Glacial Maximum (LGM), for example, is a period marked by important social and technological transformations. This climatic event was characterized by abrupt and drastic climatic changes that impacted human behavior all across westernmost Europe [@gamble2004; @sanchezgoñi2010]. The identification of a Proto-Solutrean phase in central and southern Portugal with a very distinct index fossil (the Vale Comprido point), and its direct association with the HE 2, put these regions amongst some of the most important case studies of how environmental dynamics have affected human adaptations during the last glacial. Other studies have expanded upon this notion of climatic refugia during harsh climatic events, to understand the Iberian Peninsula as a long-term eco-cultural refugia [@cascalheira2017]. Using this framework, this territory would consist of several ecological niches, consistently used through time, possibly due to the stability in the richness and variety of resources. This continuous use would then create long-term regional adaptive structures, which when correlated with the ecological niches, @cascalheira2017 have referred to as eco-cultural niches. In fact, it seems that a large number of caves and rockshelters in Iberia are multi-layered, giving validity to the aforementioned framework [@schmidt2012]. These eco-cultural niches provide an exceptional opportunity to understand long-term dynamics regarding biotic and abiotic resource exploitation since they can provide details on how human populations maintained or changed their adaptive systems when facing environmental changes, and cultural and social transformations or constraints. One of such possible Late Pleistocene eco-cultural niches to which this theoretical framework has been previously applied is the archaeological site of Vale Boi. This multi-component site is located in westernmost southern Iberia, in a region currently known as the Algarve, and comprises one of the most complete Upper Paleolithic chronocultural sequences of southern Iberia [@bicho2013; @cascalheira2013; @cascalheira2017]. Several cross-scale complex interactions have been identified, displaying resilient behaviors throughout the Upper Paleolithic maintained by their eco-cultural niche, but also adaptation behaviors motivated by niche diversity, social networks, and climatic changes [@cascalheira2017]. Some of these resilient elements are, for example, the continuous use of strategies like grease rendering and selective hunting patterns, site function, certain lithic technology patterns, and the functional specialization of lithic raw materials. The maintenance of social networks through the identification of possible long-distance lithic raw materials has also been suggested [@pereira2016]. The identification of these patterns has been reliant on the great amount of studies that have originated from the archaeological site of Vale Boi [i.e., @bicho2012; @bicho2017; @pereira2016; @manne2012]. A large portion of these studies has focused on lithic technology [@cascalheira2010; @marreiros2015; @belmiro2020], and unlike other regions of Portugal [i.e., @aubry2004; @aubry2009; @aubry2022; @pereira2022; @pereira2021; @costa2022; @jordão2022], archaeological studies focusing on raw materials, and especially chert, have been more scarce, focusing mostly on brief macroscopic results or the differentiation between possibly local and non-local sorts of raw materials. Despite the scarceness of these studies, chert played a significant role in early and later prehistory at Vale Boi [@cascalheira2010; @marreiros2009; @belmiro2020], and across most of Prehistory in Southern Portugal [@zambujo1998; @bicho2003]. As an essential part of late Pleistocene hunter-gatherer adaptations, lithic raw materials have the potential to provide insights into the adaptive strategies of those populations [@féblot-augustins1993; @kuhn1995; @mellars1996], mostly regarding land-use, technological organization, but also cultural and social interactions. In fact, the selection and procurement of raw material has been suggested as a key stage for the technological organization of hunter gatherer groups [@bamforth1997]. Thus, changes in the frequencies of raw materials within the archaeological records may provide evidence for changes in that organization, or even resilience of specific choices, all of which can reflect culturally transmitted preferences within a group of hunter gatherers [@brown1999]. Several models, both formal and informal, have shown that changes of raw material in the archaeological record may reflect different procurement strategies [@binford1985; @binford1979; @gould1985; @gouldsaggers_1985], either related to mobility strategies [@ambrose1990; @kuhn2004; @kuhn1991; @mccall2007] or changes in the availability of raw materials, possibly related to environmental change [@brantingham2003; @brown1999]. Raw materials are also closely linked to the tools produced with them -- the costs of manufacturing technology can be related to the availability of raw materials in the landscape [@torrence1983; @bousman1993; @parry1987; @mackay2011], or the characteristics of the raw materials which might relate to functionality [@mackay2008; @minichillo2006; @stout2002], through the choice of raw materials for specific technologies [@brown1999]. Furthermore, changes in the raw materials can also be related to the establishment of social networks [@whallon2006] and the horizontal transmission of preferences through trade [@brown1999]. In order to explore the aforementioned theories and models to an archaeological site, especially when trying to understand land-use and the adaptive strategies of an eco-cultural niche like what is proposed for southwesternmost Iberia, it is necessary to know the landscape and the raw material sources available in the territory. To distinguish between local and non-local raw materials, first it is necessary to understand the characteristics of the local sources and establish a comparative database which can be used for the analysis of an assemblage. This is true for most of the informal models which try to understand why raw materials change in the archaeological record. Formal models focusing on raw material use posit that it is necessary to have detailed sourcing data in order to adequately apply the models to a region or an archaeological assemblage [@surovell2009; @pop2015]. In fact, the creation of such a database is a starting point for most raw material-focused studies and has been applied all across the world, independent of the studied chronology [i.e., @luedtke1992; @bustillo2009; @driscoll2016; @terradas2017; @ekshtain2022]. In southern Portugal previous efforts have been made to create such a database. The work of @verissimo2004, albeit focusing on the occurrence of chert solely in western Algarve, provided the initial basis for comparative studies with the assemblages from some Late Pleistocene sites based on macroscopic approaches. The creation of LusoLit, a lithotheque currently hosted at the University of Algarve [@pereirab_presenting_2016], and the collection of samples from several outcrops in the region provided a new leap in the study of chert in the region. A few geological studies have also contributed to understand the availability and characteristics of chert in southern Portugal [i.e., @rocha1976; @ribeiro2005]. Nevertheless, these studies are often unpublished or comprehend answers to geological questions, which hamper the comparative use of the data with archaeological assemblages. A large portion of chert-bearing outcrops in southern Portugal remains unstudied, both by archaeologists and geologists. As such, further analysis of the overall variability, location and availability of chert nodules in the Algarve is necessary to start testing behavioural models of land-use and abiotic resource exploitation. Given the potential for a in-depth raw material study at a possible Late Pleistocene eco-cultural niche such as Vale Boi, in this study we explore the location, diversity and availability of chert raw materials in the southernmost region of Portugal -- the Algarve. Our main goal is to establish a reference for chert raw materials in an understudied region in what regards to lithic materials, that can be used for future studies addressing chert exploitation in the Algarve and elsewhere. This includes the development of a methodological approach adapted to the study area, by testing the potential of macroscopic and petrographic approaches for the characterisation of regional cherts. The data presented here is also fully integrated in an online lithotheque (LusoLit) that is now freely available and can be built upon and improved as new data becomes available.

@@

In this study we identified several sources of chert nodules in southern Portugal and characterized the regional cherts, which were of critical importance for hunter-gatherer communities during the Late Pleistocene. For this we applied a multilayered raw material analysis approach, composed of macroscopic and petrographic analyses. The results show the presence of four different chert formations, dispersed in western and eastern Algarve. Within formations, there is variability in the nodules and the outcrops. There are however identifiable macroscopic and petrographic differences between formations which allow their distinction. Although the petrographic analysis is essential to identify the fossils present in the chert, a macroscopic approach seems to be pertinent for a quick and inexpensive analysis to distinguish between cherts of the different formations. The presence of chert sources in the Algarve region, with distinguishable characteristics between formations, which may be analysed preliminarily through macroscopic approaches, shows the potential for chert raw material studies of archaeological sites in this key area. Further steps in our study will include the use of the data gathered in the present study and the completed LusoLit lithotheque to study chert use from multilayered sites with Upper Paleolithic chronologies such as Vale Boi, and participate in the discussion of human adaptations throughout the Late Pleistocene. Furthermore, future approaches include the use of geochemical methods to further characterize these cherts and the integration of the resulting data in the online database.

@@

We would like the thank the entities that funded the current research: Fundação para a Ciência e a Technologia (FCT) and ICArEHB (Interdisciplinary Center for Archaeology and the Evolution of Human Behavior). Furthermore, we want to thank Dr. Telmo Pereira and Dr. Carlos Ribeiro for allowing us to consult their previous research and for their support. We'd like to thank Jack Acres for the IT support essential for the creation of the online LusoLit database.  João Cascalheira is funded by Portuguese Ministry for Science and Technology, contract reference DL57/2016/CP1361/CT0026. Current work on chert identification in Algarve is being supported by the European Research Council through the FINISTERRA project (101045506-FINISTERRA-ERC-2021-COG) attributed to João Cascalheira.
